### 2015-07-30 Harris_2015/manuscript-materials/draft.Rmd
tighten up early methods

#### Conditional relationships and Markov networks.
Ecologists are often interested in learning about direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall") relationships.
The example with the shrubs and trees in
Figure 1 shows how the two measures can have opposite signs,
and suggests that conditional relationships are more relevant
for drawing inferences about species interactions (e.g. competition).

Markov networks define species interactions in conditional terms, and use these
conditional relationships to determine whether a given combination of co-occurring
species is consistent with their interactions.  Given a set of binary outcomes
for each species (i.e. 0 for absence and 1 for presence), the Markov network
defines the relative probability of observing a given presence-absence vector,
$\vec{y}$, as

@@

Here, $\alpha_{i}$ is an intercept term determining the amount that
the presence of species $i$ contributes
to the log-probability of $\vec{y}$; it directly controls the prevalence of
species $i$. Similarly, $\beta_{ij}$ is the amount that the co-occurrence of
species $i$ and species $j$ contributes to the log-probability; it controls
the probability that the two species will be found together (Figure 2A, Figure 2B).
Because the relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends to produce assemblages
that have many pairs of positively-associated species and relatively few pairs
of negatively-associated species (exactly as an ecologist might expect).

A major benefit of Markov networks is the fact that the conditional
relationships between species can be read directly off the matrix of $\beta$
coefficients [@murphy_machine_2012]. For example, if the coefficient linking
two mutualist species is $+2$, then---all else equal---the odds of observing
either species increase by a factor of $e^2$ when its partner is present
[@murphy_machine_2012]. In this way, the $\beta$ coefficients behave
very similarly to the coefficients in a logistic regression model
[@lee_learning_2012].

Of course, if all else is *not* equal (e.g. Figure 1,
where the presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from this
expectation. For this reason, it is important to consider how coefficients'
effects propagate through the network, as discussed below.

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a normalizing constant,
$Z$, so that that the probabilities of all possible assemblages will sum to one (bottom of
Figure 2B). Calculating $Z$ exactly, as is done in these analyses,
quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
required spirals exponentially into the billions and beyond.
Numerous techniques are available for keeping the calculations tractable,
e.g. via analytic approximations [@lee_learning_2012] or Monte Carlo
sampling [@salakhutdinov_learning_2008], but they are beyond
the scope of this paper.  Alternatively, as noted below, some common linear
and generalized linear methods can also be used as computationally efficient
approximations to the full network.

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous two sections, the values of $\alpha$ and $\beta$ were known. In
practice, however, ecologists will often need to estimate these parameters from
co-occurrence data (i.e. from a matrix of ones and zeros indicating which
species are present at which sites). When the number of species is reasonably
small, one can compute exact maximum likelihood estimates for all of the
$\alpha$ and $\beta$ coefficients given a presence-absence matrix by optimizing
$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the ones
considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
(i.e. it matches the sufficient statistics of the model distribution to the
sufficient statistics of the data). For the analyses in this paper, I used the
*rosalia* package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to define the objective function and gradient in R code
and then find the optimum.

>>>>

#### Conditional relationships and Markov networks.
Ecologists are often interested in learning about direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall")
relationships. The example with the shrubs and trees in Figure 1 shows how the
two measures can have opposite signs, and suggests that conditional
relationships are more closely related to the underlying ecological interactions.

Markov networks provide a method for translating back and forth between the
conditional relationships among species and the kinds of species assemblages
that these relationships produce. In this section, I show how a set of
conditional relationships can be used to determine the kinds of species
assemblages that can occur. Methods for estimating conditional relationships
from presence-absence data are discussed in the next section.

Given a set of binary outcomes for each species (i.e. 0 for absence and 1 for
presence), a Markov network defines the relative probability of observing a
given presence-absence vector, $\vec{y}$, as

@@

Here, $\alpha_{i}$ is an intercept term determining the amount that the presence
of species $i$ contributes to the log-probability of $\vec{y}$; it directly
controls the prevalence of species $i$. Similarly, $\beta_{ij}$ is the amount
that the co-occurrence of species $i$ and species $j$ contributes to the
log-probability; it controls the conditional relationship between two species,
i.e. the probability that they will be found together, after controlling for the
other species in the network (Figure 2A, Figure 2B). Because the relative
probability of a presence-absence vector increases when positively-associated
species co-occur and decreases when negatively-associated species co-occur, the
model tends to produce assemblages that have many pairs of positively-associated
species and relatively few pairs of negatively-associated species (exactly as an
ecologist might expect).

Of course, if all else is *not* equal (e.g. Figure 1,
where the presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from this
expectation. Predicting the marginal relationships between species from their
conditional interactions entails summing over the different possible assemblages
(Figure 2B), and the number of possible assemblages can be intractably large.
Numerous techniques are available for keeping the calculations feasible,
e.g. via analytic approximations [@lee_learning_2012] or Monte Carlo
sampling [@salakhutdinov_learning_2008], but they are beyond
the scope of this paper.  Alternatively, as noted below, some common linear
and generalized linear methods can also be used as computationally efficient
approximations to the full network.

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a matrix of ones and zeros
indicating which species are present at which sites). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
(i.e. it matches the sufficient statistics of the model distribution to the
sufficient statistics of the data). For the analyses in this paper, I used the
*rosalia* package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to define the objective function and gradient in R code
and then find the optimum.
